{\bf {\Large Four decades of water quality change in the upper San Francisco Estuary}}

### 2016-12-12 sftrends_manu.Rnw
minor edits to methods

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 to 2012.  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \citep{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not signifcantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.  
\subsection{Analysis method and application}

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.  The \ac{wrtds} model is a moving window regression that fits a unique set of parameters at each observation point in the time series.  A unique set of weights is used for each regression to control the relevance of observations used to fit the model to the observation at the center of the window. The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.  The final vector used to fit the model at each point weights observations more similar to the center of the window with more importance.  The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.  As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series.  The original \ac{wrtds} method is described in more detail in \citep{Hirsch10}.  The \ac{wrtds} model applied to the Delta time series followed methods in \citet{Beck15}, which were based on a tidal adaptation of the original method.  The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.  Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme evens.  The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.  All analyses used the WRTDStidal package for R \citep{Beck16b,RDCT16}

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.  Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}). Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends. A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.  Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2014, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter). These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16}.

Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.  The results were interpreted relative to those in \citet{Jabusch16} that were based on observed data to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.  The analysis was also similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression. Functions in the EnvStats package in R were used for the seasonal Kendall tests \citep{Millard13}.

@@

Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.  The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends on decadal time scales.  The selected case studies focused on \begin{inparaenum}[1\upshape)]
\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
\item effects of wastewater treatment upgrades upstream of P8, and
\item effects of biological invasion on nutrient dynamics in Suisun Bay.
\end{inparaenum}
Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.

@@

The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \citep{Cohen98}.  The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \citep{Carlton90,Nichols90,Werner93,Crauder16}.  In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.  Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \citep{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}.  Abundance of clams is linked to inflow rates such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.

We hypothesized that \ac{wrtds} models applied to water quality obsevations in the upper estuary would emphasize \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
\end{inparaenum}
The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.

@@

<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.  Changes are based on seasonal Kendall tests of flow-normalized results within each time period. Station names are shown in the top left panel. Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trnds} for actual values. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 8, cache = T, out.width = '\\textwidth'>>=

@@

grid.arrange(
  pA, pB, pC, pD,
  heights = c(0.9, 1.3, 1, 1.2),
  ncol = 1
  )
@

>>>>

Something about flow

@@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \citep{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \citep[references in][]{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at nine discrete sampling stations from 1976 through 2013.  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

Daily flow estimates for the Delta region were obtained from the Dayflow software program that provides estimates of average Delta outflow \citep{IEP16}.  Because of the complexity of water inflow, exports, and outflows from the Delta, the Dayflow program combines observations with estimates based on mass balance to reconstruct historical and daily flow estimates.  The \ac{wrtds} models described below require a matched flow record with the appropriate station to evaluate nutrient trends. Given the complexity of inflows and connectivity of the system, only the inflow estimates from the Sacramento and San Joaquin rivers were used as measures of freshwater influence at each station.  Initial analyses indicated that model fit was not significantly improved with flow estimates from locations closer to each station, nor was model fit improved using lagged times series.  As such, the Sacramento daily flow time series was used to acount for flow effects at C3, D19, D26, and D28, and the San Joaquin time series was used for C10 and P8.  The salinity observations at D4, D6, and D7 in Suisun Bay were used as a more appropriate measure of variation in freshwater balance given the stronger tidal influence at these stations.  Salinity has been used a tracer of freshwater influence for the application of WRTDS models in tidal waters \citep{Beck15}.

\subsection{Analysis method and application}

@@

\noindent where $N$ is one of three nitrogen analytes, time $t$ is a continuous variable as decimal time to capture the annual or seasonal trend, and $Q$ is the flow variable (either flow or salinity depending on station).  The seasonal trend is modelled as a sinusoidal component to capture periodicity between years.  The \ac{wrtds} model is a moving window regression that fits a unique set of parameters at each observation point in the time series.  A unique set of weights is used for each regression to control the relevance of observations used to fit the model to the observation at the center of the window. The weights are based on a scaled Euclidean distance to estimate the differences of all points from the center in relation to annual time, season, and flow.  The final vector used to fit the model at each point weights observations more similar to the center of the window with more importance.  The complete model for the time series contains a parameter set for every time step that considers the unique context of the data.  As such, predictions from \ac{wrtds} are more precise than those from more conventional models that fit a single parameter set to the entire time series \citep{Moyer12,Beck15}.  The original \ac{wrtds} method is described in more detail in \citep{Hirsch10}.  The \ac{wrtds} model applied to the Delta time series followed methods in \citet{Beck15}, which were based on a tidal adaptation of the original method.  The \ac{wrtds} models were fit to describe the conditional mean response using a weighted Tobit model for left-censored data \citep{Tobin58}.  Previous adaptations of \ac{wrtds} to tidal waters have used quantile regression to describe trends in the conditional quantiles, such as changes in the frequency of occurrence of extreme evens.  The application to the Delta data focused only on the conditional mean models to establish a baseline response which has not been previously quantified.  All analyses used the WRTDStidal package for R \citep{Beck16b,RDCT16}

A hallmark of the \ac{wrtds} approach is the description of flow-normalized trends that are independent of variation from freshwater inflows.  Flow-normalized trends have value for the interpretation of changes that are potentially caused by drivers other than  flow, such as \ac{wwtp} upgrades or phytoplankton grazing by benthic filter-feeders (\cref{fig:schematic}, \citealt{Beck15}). Although variation in nutrients is caused by the combined effects of several variables acting at different temporal and spatial scales, flow-normalization provides a basis for further exploration by removing a critical confounding variable that could affect the interpretation of trends. A flow-normalized value is the average of predictions at a given observation using all flow values that are expected to occur for the relevant month across years in the record.  Flow-normalized trends for each analyte at each station were used to describe long-term changes in different annual and seasonal periods.  Specifically, flow-normalized trends in each analyte were summarized as both medians and percent changes from the beginning to end of annual groupings from 1976-1995 and 1996-2013, and seasonal groupings of March-April-May (spring), June-July-August (summer), September-October-November (fall), and December-January-February (winter) within each annual grouping. These annual and seasonal groupings were chosen for continuity with similar comparisons reported in \citet{Jabusch16} and as approximate twenty year midway points in the time series.

Trends within each annual and seasonal grouping were based on seasonal Kendall tests of the flow-normalized predictions. This test is a modification of the non-parametric Kendall test that accounts for variation across seasons in the response variable \citep{Hirsch82}.  Results from the test can be used to evaluate the direction, magnitude, and significance of a monotonic change within the period of observation.  The estimated rate of change per year is also returned as the Theil-Sen slope and was interpreted as the percent change per year when divided by the median value of the response variable in the period of observation.  Trends within annual groupings were based on all monthly observations within relevant years, whereas seasonal groupings were based only on the relevant months across all years in the data set.  Seasonal Kendall tests weere also used to describe trends in the model predictions for the observed data.  These trends were compared with those based on the flow-normalized trends to evaluate the improved ability of \ac{wrtds} to describe trends that are independent of flow.  The analysis was similar to that in \citet{Jassby08} that evaluated trends after accounting for river inflow using a more generic form of locally weighted regression. Functions in the EnvStats package in R were used for the seasonal Kendall tests \citep{Millard13}.

@@

Three stations were chosen for closer evaluation to demonstrate use of \ac{wrtds} to develop a more comprehensive description of decadal trends in the Delta.  The stations were chosen to address ecological and management-based questions that have relevance outside of the region, having importance for the understanding of estuarine processes that influence eutrophication trends over several years.  The selected case studies focused on \begin{inparaenum}[1\upshape)]
\item nitrogen trends and flow-effects of the San Joaquin river at C10, 
\item effects of wastewater treatment upgrades upstream of P8, and
\item effects of biological invasion on nutrient dynamics in Suisun Bay.
\end{inparaenum}
Each case study is built around hypotheses that results from \ac{wrtds} models were expected to support, both as a general description and for additional testing with alternative methods.

@@

The \acl{sfe} is considered one of the most invaded ecosystems in the world with an estimated 234 exotic species by the turn of the century, half of which have been reported after 1965 \citep{Cohen98}.  The invasion of benthic grazers as ecosystem engineers is one of the more notable events that has been characterized by dramatic shifts in primary production of the Bay's trophic network \citep{Carlton90,Nichols90,Werner93,Crauder16}.  In particular, invasion of the upper estuary by the Asian clam \textit{Potamocorbula amurensis} in 1986 caused dramatic changes in phytoplankton abundance and species composition with increased grazing.  Reduction in phytoplankton biomass has altered trophic networks in the Bay and is considered a primary mechanism in the decline of the protected delta smelt and other important fisheries \citep{Feyrer03,MacNally10}.  Changes in the physical environment have also occurred with the most notable effect being increased water clarity following a reduction of phytoplankton \citep{MacNally10}. The clams are halophilic such that drought years are generally correlated with an increase in biomass and further upstream invasion of the species \citep{Parchaso02,Cloern12}.

We hypothesized that \ac{wrtds} models applied to water quality observations in the upper estuary would show \begin{inparaenum}[1\upshape)]
\item a decline in annual, flow-normalized chlorophyll concentrations over time coincident with an increase in abundance of invaders, 
\item changes in ratios of limiting nutrients (nitrogen, SiO$_2$) suggesting different uptake rates by grazers with a shift in community composition, and 
\item seasonal shifts in limiting nutrients based on changes in community composition and relative abundances with seasonal succession. 
\end{inparaenum}
The application of \ac{wrtds} to water quality observations at station D7 in Suisun Bay and comparison with clam abundance and biomass data from \cite{Crauder16} was expected to reveal the competing effects of inflow on phytoplankton and benthic grazers.

@@

<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.  Changes are based on seasonal Kendall tests of flow-normalized results within each time period. Station names are shown in the top left panel. Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 8, cache = T, out.width = '\\textwidth'>>=

@@

grid.arrange(
  pA, pB, pC, pD,
  heights = c(0.9, 1.3, 1, 1.2),
  ncol = 1
  )
